Formal Languages and Automatas

### 2024-08-15 paper/main.tex
at kleene proof

\documentclass[12pt,letterpaper]{article} 
\usepackage[letterpaper,margin=1in]{geometry} 
\usepackage{mathptmx}
\usepackage[singlespacing]{setspace}
\usepackage{fancyhdr}
\setlength{\headheight}{14.49998pt}
\usepackage{relsize}
\usepackage[bottom]{footmisc}
\usepackage{tabularx}
\setlength{\parindent}{2em}
\usepackage{url}
\usepackage{hyperref}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{expex}
\usepackage{amsfonts} 
\usepackage{tikz}
\bibpunct[:]{(}{)}{;}{a}{}{,}
\usepackage{bibentry}
\usepackage{acronym}
\usepackage{multicol}
\author{Ruben Triwari}
\usepackage{scrhack} 
\usepackage{microtype}
\usepackage[acronym,xindy,toc]{glossaries}
\makeglossaries
\loadglsentries{glossary.tex}
\usepackage{ragged2e}
\babelhyphenation[english]{
an-oth-er
ex-am-ple
}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}[section]
\usepackage{dcolumn}
\newcolumntype{2}{D{.}{}{2.0}}
\usetikzlibrary{
  automata, positioning,
  arrows, matrix, decorations.pathreplacing,
  shapes.geometric
}

@@

First of all, we need to define some simple structures and ideas,

@@

the previous definition, becaus we can always find an isomorphism between

@@

we can always number a finite amount of objects. The above definition

@@

  $\Sigma$. The empty word which contains no characters, 
  is denoted by $\epsilon$

@@

  sometimes just written as $wv$. With powers:
    w^n = \underbrace{w\cdot w \cdot \dots w}_{\text{n times}}
    \text{ where } n \in \mathbb{N}
  \[ w^0 = \epsilon.\]

@@

  \[ L \cdot V := \{ l \cdot v | l \in L \lor v \in V \}\]

@@

      L^{i+1} & := \{ l_i\cdot l | l_i \in L^i \lor l \in L\}

@@

  written as:

@@

    (L,w) \in \mathbb{B} = \{0,1\}

@@

@@

The sum is iterating over all possible words $w \in \Sigma^*$ and 
$(L,w)$ is an indicating function, which is one if $w$ should be in the
language and ohterwise zero. Thus all words with an one infront are in 
the language.\\

@@

  Let $\Sigma$ be an fixed alphabet $L, V$ formal languages,
  L and V can be written

@@

    L \cdot V := \sum_{w \in \Sigma^*} ((L,u) \cdot (V,v))w
    \verb|,|\hspace{0.5cm} \verb|where | w = uv.

@@

    L^{i+1} &:= \sum_{w \in \Sigma^*} ((L^i,u)\cdot (L,v))w, 
    \hspace{0.5cm} \text{where} \hspace{0.5cm} w = uv.

@@

  Frthermore, we need $H(i,j) = \{ \alpha \in \Sigma : \delta(s_i, \alpha) = s_j\}$
  and two helper functions:

@@

    w: \Sigma^\times \to \Sigma^*, 
    \delta': \mathbb{N} \times \mathbb{N} \to \Sigma^*

@@

    w(\{a_1,a_2,\dots, a_n\}) = a_1a_2\dots a_n

@@

        w(H(i,j)), & \text{if} \hspace{0.2cm} \exists \alpha \in \Sigma: 

@@

    $A \in (\Sigma^*)^{n \times n}$

@@

  We can denote a state transition $s_i \to s_j$ with $s_iA_{ij} = s_j$.

@@

    A^* = E_n + A + A^2 + A^3 \dots  = \sum_{i \in \mathbb{N}_0} A^i.

@@

the state set, since as mentioned above, there always exist an isomorphism

@@

\begin{example}
  Here is a simple automata depicted, where the circles are states
  , the arrows are transitions, the start is indicating the initial 
  state, and the double lined circle is the accepting state.
  \[ 
    M_\text{tuple} = (\{s_1, s_2\}, \{a,b,c\}, \delta_0, s_1, \{s_2\}),
    M_\text{matrix} = (\{a,b,c\}, A, s_1, \{ s_2 \})
  \]\\
  where
  \[
    \delta_0(s_1, a) = s_1, \hspace{0.2cm} 
    \delta_0(s_1, b) = s_2, \hspace{0.2cm} \delta_0(s_2,c) = s_2
  \]
  and
  \[ 
    A = 
    \begin{pmatrix}
      a & b \\
      0 & c 
    \end{pmatrix}.
  \]
  \begin{center}
    \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
      \node[state,initial] (q1) at (0,0) {$s_1$};
      \node[state,accepting] (q2) at (3,0) {$s_2$};
      \path[->] (q1) edge[loop above] node{a} (q1)
      (q2) edge[loop above] node{c} (q2)
      (q1) edge[bend left] node{b} (q2);
    \end{tikzpicture}
  \end{center}
  The resulting accepting formal language is:
  \[ \text{Set represntation: } L = \{a^nbc^n : n \in \mathbb{N}\}\]
  \[
    \text{Formal power series represntation: } L = a^*\cdot b\cdot c^* 
  = a^*bc^*
\]
\end{example}
% Define Formal languages (Sets and Power sereies) 
% and automatas (Tuple and Matrix)
\section{Formal Languages: Sets vs. Formal Power Series}
% translate one in another
% Formal power series are an algebra over Boolean algebra
% U could define the same algebra for sets with set operations
% But u could replace the boolean algebra by any abritary ring
% Which makes formal power series more powerful cuz u can abstract even more
\section{Maximum Gap Lemma}
\section{Pumping Lemma}
\section{Not all Formal languages are regular}
\section{Pumping Lemma vs. Maximum Gap Lemma}
\section{Kleene's theorem (Power series, Matrix Automata)}
\section{Kleene's theorem (Set, Tuple)}
\section{Kleene's theorem vs. Kleene's theorem}
\section{Kleene Schützenberg theorem and weigthed automatas}
\section{Conclusion}

>>>>

\documentclass[12pt,letterpaper]{article} 
\usepackage[letterpaper,margin=1in]{geometry} 
\usepackage{mathptmx}
\usepackage[singlespacing]{setspace}
\usepackage{fancyhdr}
\setlength{\headheight}{14.49998pt}
\usepackage{relsize}
\usepackage[bottom]{footmisc}
\usepackage{tabularx}
\setlength{\parindent}{2em}
\usepackage{url}
\usepackage{hyperref}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{expex}
\usepackage{amsfonts} 
\usepackage{tikz}
\bibpunct[:]{(}{)}{;}{a}{}{,}
\usepackage{bibentry}
\usepackage{acronym}
\usepackage{multicol}
\author{Ruben Triwari}
\usepackage{scrhack} 
\usepackage{microtype}
\usepackage[acronym,xindy,toc]{glossaries}
\makeglossaries
\loadglsentries{glossary.tex}
\usepackage{ragged2e}
\babelhyphenation[english]{
an-oth-er
ex-am-ple
}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}[section]
\newenvironment{proof}{\paragraph{Proof:}}{\hfill$\square$}
\usepackage{dcolumn}
\newcolumntype{2}{D{.}{}{2.0}}
\usetikzlibrary{
  automata, positioning,
  arrows, matrix, decorations.pathreplacing,
  shapes.geometric
}
\newcommand{\fps}[1] {
\mathbb{#1}\langle \langle \Sigma^* \rangle \rangle
}

@@

First of all, we need to define some structures and ideas,

@@

the previous definition, becaus we can always find a one-to-one map between

@@

we can always enumerate a finite amount of objects. The above definition

@@

  $\Sigma$. 
    w = (a_1,a_2, \dots, a_n) \in \Sigma^n
  We will denote it by:
    w = a_1a_2\dots a_n
  The empty word which contains no characters, 
  and is denoted by $\epsilon$.
  %book schützenbert

@@

  sometimes just written as $wv$. With powers recursively defined by:
    w^0 &:= \epsilon\\
    w^1 &:= w\\
    w^{n+1} &:= w^n\cdot w

@@

  \[ L \cdot V := \{ l \cdot v | l \in L \land v \in V \}\]

@@

      L^{i+1} & := \{ l_i\cdot l | l_i \in L^i \land l \in L\}

@@

  written as formal power series $L$:
    L: \Sigma^* \to \mathbb{B}
    \mathbb{B} = \{0, 1\}

@@

    (L,w) \in \mathbb{B}.

@@

  {\bf Note:} The set of formal power series is denoted by 
  $\mathbb{B}\langle \langle \Sigma^* \rangle \rangle 
    = \{ \Sigma^* \to \mathbb{B}\}$\\
  {\bf Notation:} If the language consist only out of one 
  character or the empty word $a \in \Sigma \cup \{\epsilon\}$, 
  then the formal power series would be: 
    L = h \in \fps{B}
    h: \Sigma^* \to \mathbb{B}
    h(w) = 
      1 & \text{if   } w=a \\
      0 & \text{otherwise}
  In this case we just write $L = a$.
    %schützenberg

@@

This looks like a sum, we call it series, but it is actually a map from 
words to zero or one. The sum is a way to write a function definition: Infront
of every word $w \in \Sigma^*$ in the series, there is a $x \in \mathbb {B}$,
which is zero or one. So we can interpret every element in the series as an 
assignment: $w \mapsto x$.
All words that map to one, are in the language, the others are not. So we can write a formal language
as a function $\Sigma^* \to \mathbb{B}$.\\

@@

  Let $\Sigma$ be an fixed alphabet and $L, V \in \mathbb{B}\langle \langle
  \Sigma^* \rangle \rangle$, where

@@

    L \cdot V := \sum_{\substack{u,v \in \Sigma^* \\ w = uv }} 
      ((L,u) \cdot (V,v))w

@@

    L^{i+1} &:= \sum_{\substack{u,v \in \Sigma^* \\ w = uv}}
    ((L^i,u)\cdot (L,v))w, 

@@

  Frthermore, we need a helper function:

@@

    \delta': \mathbb{N} \times \mathbb{N} \to \fps{B}

@@

@@

        \alpha, & \text{if} \hspace{0.2cm} \exists \alpha \in \Sigma: 

@@

  $A \in \fps{B}^{n \times n}$

@@

  We can denote a state transition $s_i \to s_j$ with $s_iA_{ij} = s_j$.\\
  {\bf Note:} We definined $A \in \fps{B}^{n \times n}$ and not 
  $A \in \Sigma^{n \times n}$, because otherwise the defintion is not closed
  under matrix multiplication, nor matrix addition.\\

@@

    A^* = E_n + A + A^2 + A^3 \dots  = \sum_{i \in \mathbb{N}_0} A^i
  \[ E_n \in \fps{B}^{n \times n}\]
    (E_n)_{ij} := 
      \epsilon & \text{if } \hspace{0.2cm} i=j \\
      0        & \text{otherwise}
    \end{cases} \in \fps{B}

@@

the state set, since as mentioned above, there always exist a one-to-one map

@@

  Here is a simple automata depicted, where the circles are states
  , the arrows are transitions, the start is indicating the initial 
  state, and the double lined circle is the accepting state.
    M_\text{tuple} = (\{s_1, s_2\}, \{a,b,c\}, \delta_0, s_1, \{s_2\}),
    M_\text{matrix} = (\{a,b,c\}, A, s_1, \{ s_2 \})
  \]\\
    \delta_0(s_1, a) = s_1, \hspace{0.2cm} 
    \delta_0(s_1, b) = s_2, \hspace{0.2cm} \delta_0(s_2,c) = s_2
      a & b \\
      0 & c 
    \end{pmatrix} \in \fps{B}^{2\times 2}.
    \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
      \node[state,initial] (q1) at (0,0) {$s_1$};
      \node[state,accepting] (q2) at (3,0) {$s_2$};
      \path[->] (q1) edge[loop above] node{a} (q1)
      (q2) edge[loop above] node{c} (q2)
      (q1) edge[bend left] node{b} (q2);
  The resulting accepting formal language is:
  \[ \text{Set represntation: } L = \{a^nbc^n : n \in \mathbb{N}\}\]
    \text{Formal power series represntation: } L = a^*\cdot b\cdot c^* 
  = a^*bc^*
% Define Formal languages (Sets and Power sereies) 
% and automatas (Tuple and Matrix)
\section{Formal Languages: Sets vs. Formal Power Series}
In this chapter we will discuss the obvious diffrences and 
simillarities of both representations. First we will give intuition,
why one might use a seires to represent a function definition.
Secondly, we will map the formal power series version into the set
version and the other way around, to get a better understanding of both.
Lastly, we will discuss why formal power series are more powerful
than just sets.\\
The intuition behind the notation of formal power series is quite simple.
The operation defined on the formal power series, which are maps, 
are really close to a series. For addition, formal power series and 
normal series are quite simillar, namely defined pointwise.\\
In the following are $L,V \in \fps{B}$ and $A,B$ series.
  L + V = \sum_{w \in \Sigma^*} ((L,w) + (V,w))w \hspace{0.3cm}
  \leftrightarrow \hspace{0.3cm}
  A + B = \sum_{n \in \mathbb{N}} (a_n + b_n) x^n
The multiplication defined on formal power series are again really close
to the cauchy product of series.
In the following are $L,V \in \fps{B}$ and $A,B$ series.
  L \cdot V = \sum_{\substack{u,v \in \Sigma^* \\ w=uv}} 
  ((L,u) \cdot (V,v))w
  \hspace{0.3cm} \leftrightarrow \hspace{0.3cm}
  A \cdot B = 
  \sum_{\substack{i,j \in \mathbb{N} \\ n = i + j}} (a_i \cdot b_j) x^n
Consequently, it is quite handy to use a formal power series as function 
definition, because the operations seem natural with series.
Despite the notation being quite confusing at first glance.\\
%Maybe subsection
Next we will show that bot representations of a formal language are 
equivalent, by mapping one to another.
To be sure that both representation are actually equivalent, we need to
show there exist an bijective map, between formal power series and the
set of words. This map needs to preserve the structure, that means the
order of function application and operator application must be irrelevant.
Such maps are called isomporphism and structures for that a isomorphism 
exist are called isomporphic to each other. Thus we want tho show that
both representation are isomorphic.
\begin{proof}
  Let $\Sigma$ be an fixed alphabet, then we claim that:
    \mathbb{B}\langle \langle \Sigma^* \rangle \rangle
    \cong \mathcal{P}(\Sigma^*)\]
  In order to show those are isomorphic, we provide a map $\phi: 
  \mathbb{B}\langle \langle \Sigma^* \rangle \rangle \to \mathcal{P}(\Sigma^*)$ 
  and its inverse $\phi^{-1}: \mathcal{P}(\Sigma^*) \to \fps{B}$ and
  show that $\phi$ is bijective and preservers structure, formally:\\
  Let $L,V \in \fps{B}$, then for $\phi$ must hold:
  \begin{align}
    \phi(L + V) &= \phi(L) \cup \phi(V) \\
    \phi(L \cdot V) &= \phi(L) \cdot \phi(V)\\
    \phi(L^*) &= \phi(L)^*
  \end{align}
  We define $\phi$ and its inverse as follows:
    \phi(L) = \{ w \in \Sigma^* : (L,w) = 1\}
    \phi(L)^{-1} = h_{L}
  \[ h_L: \Sigma^* \to \fps{B}\]
    h_L(w) = 
      1, & \text{if} \hspace{0.2cm} w \in L \\
    \end{cases}
  To prove that $\phi$ is bijective, it must hold
      \phi^{-1}(\phi(L)) = L
      \phi(\phi^{-1}(L)) = L.
    First we show $\phi^{-1}(\phi(L)) = L$\\
    Let $\ L \in \fps{B}$ abitrary but fixed, with 
    \[ L =\sum_{w \in \Sigma^*} (L,w)w. \]
    Then
      \phi^{-1}(\phi(L)) = \phi^{-1}(\{ w \in \Sigma^* : (L,w) =1 \})
       &= 
        1 & \text{if} \hspace{0.2cm} w \in \{ w \in \Sigma^* : (L,w) =1 \}\\
        0 & \text{otherwise}
       \end{cases}\\
       &= 
        1 & \text{if} \hspace{0.2cm} (L,w) = 1\\
        0 & \text{otherwise}
       \end{cases}\\
       &= 
       \sum_{w \in \Sigma^*} (L,w)w = L.
    Next we need to show that $\phi(\phi^{-1}(L)) = L.$\\
    Let $\ L \in \mathcal{P}(\Sigma^*)$ abitrary but fixed, then
      \phi(\phi^{-1}(L)) = \phi(h_L) = \{w \in \Sigma^* : (h_L,w) = 1 \}
      = \{ w \in \Sigma^* : h_L(w) = 1\} = \{ w \in \Sigma^* : w \in L\} = L.
    Consequently, $\phi$ is a one-to-one correspondence between formal power
    series and set of words.
    We are not done yet, we still need to prove that $\phi$ preserves
    structure, thus we need to show $(1), (2),$ and $(3)$. Starting with
    number one:
    \[ \text{\bf (1)} \hspace{0.3cm} \phi(L+V) = \phi(L) \cup \phi(V) \]
    In the following let $L,V \in \fps{B}$ arbitrary but fixed. Then
      \phi(L+V) &= \{ w \in \Sigma^* : (L,w) + (V,w) = 1 \} \\
        &= \{ w \in \Sigma^* : (L,w) = 1 \lor (V,w) = 1 \} \\
        &= \{ w \in \Sigma^* : w \in \{ x \in \Sigma^* : (L,x) = 1 \}
        \lor w \in \{ x \in \Sigma^* :(V,x) = 1 \} \} \\
        &= \{ w \in \Sigma^* : w \in \phi(L) \lor w \in \phi(V) \} 
        = \phi(L) \cup \phi(V).
    Continuing with two:
      \text{\bf (2)} \hspace{0.3cm} 
      \phi(L\cdot V) = \phi(L) \cdot \phi(V) 
    In the following let $L,V \in \fps{B}$ arbitrary but fixed. Then
      \phi(L\cdot V) &= \{ uv \in \Sigma^* : (L,u) \cdot (V,v) = 1\}\\
        &= \{ uv \in \Sigma^* : (L,u) = 1 \land (V,v) = 1\} \\
        &= \{ uv \in \Sigma^* : u \in \{ w \in \Sigma^* : (L,w) = 1\}
          \land v \in \{ w \in \Sigma^* : (V,w) = 1\}
        \} \\
        &= \{ uv \in \Sigma^* : u \in \phi(L)
          \land v \in \phi(V)
        \} = \phi(L) \cdot \phi(V).
    Last but not least, three:
      \text{\bf (3)} \hspace{0.3cm} 
      \phi(L^*) = \phi(L)^*
    In the following let $L \in \fps{B}$ arbitrary but fixed. Then
      \phi(L^*) = \phi(\sum_{i\in \mathbb{N}_0} L^i) 
      \stackrel{\text{\bf (1)}}{=} \sum_{i\in \mathbb{N}_0} \phi(L^i)
      \stackrel{\text{\bf (2)}}{=} \sum_{i\in \mathbb{N}_0} \phi(L)^i
      = \phi(L)^*
    Thus $\phi$ is an isomorphism and our claim follows.
\end{proof}
\\\\
Now we have shown that both definition are equivalent, but why should
one use the less natural formal power series definition? The anwser
to this, as so often in mathematics, is {\bf abstraction}. Finite automata
are accepting formal languages, thus we can say they generate a formal
language. When abstracting from finite automata's to weighted automata's
which have a weight or cost between each transition, the generated
object is not longer a language, but a map that assigns a weight or
cost to each word. Generally, this weight or cost can be seen as a 
semiring (a Ring that does not have necessarily an additive inverse),
which are objects that behave like numbers, we will denote an arbitrary
semiring with $\mathbb{S}$. Those maps, that get accepted by an weighted
automata, have the signature $L: \Sigma^* \to \mathbb{S}$ and are 
indeed a formal power series over $\mathbb{S}$, consequently $L \in \fps{S}$.
To Summarize, finite automata generate formal power series over $\mathbb{B}$
and weighted automata generate formal power series over $\mathbb{S}$. So 
formal power series are in fact an abstraction of formal languages.
\section{Kleene's theorem (Power series, Matrix Automata)}
In this chapter we will examine, what kind of languages get accepted 
by an finite automata. In order to do that, we will prove that all
languages that get accepted by an finite automata are rational and every
rational language can be generated by an finite automata (kleene's theorem).
A rational language is defined inductively.
  Let $\Sigma$ be an fixed alphabet, then a rational language 
  is defined as follows:
    \item A language $L = a \in \fps{B}$ with a single character 
      $a \in \Sigma$ is said to be rational.
    \item The empty word $\epsilon$ is rational.
    \item The sum of two rational Languages $L_1, L_2 \in \fps{B}$
      are again rational $L = L_1 + L_2$.
    \item The product of two rational Languages $L_1, L_2 \in \fps{B}$
      are again rational $L = L_1 \cdot L_2$.
    \item The kleene star of a rational language $L \in \fps{B}$ 
        is again rational $V = L^*$.
So we start with just one character and then building with our operations
step by step  a language. To show kleene's theorem we have to prove two
direction. First we will prove that finite automata only accept rational 
languages. Afterwards, we will prove that every rational language is 
accepted by an finite automata. In an effort to do that, we need to 
understand certain properties of the Matrix encoding of finite automatas.
Raising the power of two of a Matrix is defined by:\\
Let $A \in \fps{B}^{n\times n}$ be an arbitrary matrix, then
  (A^2)_{ij} = \sum_{k=1}^{n} A_{ik}A_{kj}.
As mentioned before, this matrix is a transistion function of an finite
automata, where we can look up, what character carry one state in another.
In $A_{ij}$ is the character that carry $i$ to $j$, thus multiplying
$A$ by itself, gives us a new Matrix with all words with length two,
given as input and $i$ is the starting node would
carry $i$ to $j$ in two steps. In the formula above we see that we
are trying to go throug all nodes $k \in \{1, 2, 3, \dots n\}$
to end up in $j$ and then uniting them together with addition.\\
Generally, we can have a matrix $A \in \fps{B}^{n \times n}$
risen to the power of $n \in \mathbb{N}$:
  (A^n)_{ij} = \sum_{k_1=1}^{n}
    \sum_{k_2=1}^{n}\dots \sum_{k_{n-1}=1}^n 
    a_{ik_1}a_{k_1k_2}a_{k_2k_3}\dots a_{k_{n-1}j} \\
This is not beatiful at all, but the idea stays exactly the same: Now
we have in $(A^n)_{ij}$ all words that carry $i$ to $j$ in n steps,
brute forcing throug all possible paths. Now to get all possible paths
of arbitrary length we just add all together:
  E_n + A + A^2 \dots = \sum_{i \in \mathbb{N}} A^i = A^*
Yielding, that in $(A^*)_{ij}$ are all possible words, carrying
the automata from $i$ to $j$. If $i$ is the starting node and $j$
the accepting node, the accepting language of the automata is just
$(A^*)_{ij}$. That means, $A^*$ stores all accepting languages of 
an automata, with transition matrix $A$.
With those important properties, we can now move on to the first prove.
\begin{proof}
    {\bf claim: All finite automata only accept rational languages.}
  Let $A \in \fps{B}^{n \times n}$ be a transition function of an
  arbitrary finite automata. $A$ is rational, because the transition
  function consists only out of characters. Since $A^*$ stores every accepting
  language of that finite autotamata, for every possible initial
  state and accepting state, we just need to prove that all entries
  in $A^*$ are rational languages.
  We will prove this statement via induction over the number of 
  states, namely the dimension of the square matrix $n \in \mathbb{N}$.\\
  {\bf Base case $n=1$}:\\
  Thus
    A \in \fps{B}^{1 \times 1} = \fps{B}.
  We know that $A \in \fps{B}$ is rational, but that also means
  that $A^*$ is per definition rational.

\end{proof}

\section{Maximum Gap Lemma}
\section{Not all Formal languages are regular}
\section{Kleene Schützenberg theorem and weigthed automatas}
\section{Conclusion}
